---
title: "Gov 50 Final Project"
author: Robert Zipper
description: "The Relationship Between Poverty Rate and Economic Ideology in Congress"
output:
  distill::distill_article:
    self_contained: false
---


## Project Thoughts

In American politics, it is common knowledge that the two main political parties have divergent economic ideologies, specifically regarding redistribution of income by the federal government. The Democratic Party has generally been more receptive than the Republican Party to federal income-redistribution programs, including poverty relief programs. 

However, voters do not only vote according to their economic interests -- increasingly, they have also been voting according to their cultural and other related ideological interests. Therefore, many voters in high-poverty regions of the U.S. do not vote for Democratic candidates who are more inclined to support economic policies (e.g., poverty relief programs) that would benefit these voters. As a result, their representatives may not be supporting the enactment of policies that would reduce poverty in these communities. 

For this project, I would like to examine the extent to which a U.S. congressional district's poverty rate is correlated with the economic voting ideology of that district's representative in Congress. I plan to use the DW-NOMINATE (DW-N) scores of members of Congress as a way to measure their economic ideology on the left-right spectrum, based on their voting records. A lower DW-N score suggests that a representative is more economically liberal and thus more likely to support federal anti-poverty programs. I expect to see a positive correlation, but I do not think it will be a strong one, given the geographical and partisan heterogeneity of high-poverty regions across the United States.


## Project Proposal

Do members of Congress serving districts with high rates of poverty support more progressive economic policies? In this study, I plan to examine the extent to which a U.S. congressional district's poverty rate correlates with its U.S. Representative's economic ideology, as measured by the representative's voting record in Congress. It is reasonable to assume that members of Congress support policies that are in the best economic interests of their constituents. Therefore, I hypothesize that members of Congress representing districts with high rates of poverty are more likely to support progressive economic policies because such policies tend to benefit poorer Americans. For example, progressive economic ideology is associated with greater spending on anti-poverty programs, which, of course, disproportionately benefit those living in high-poverty regions. My sample is comprised of the poverty rates for each U.S. congressional district compiled by the Food Research and Action Center, based on the 2017 American Community Survey. My unit of analysis is a congressional district, and my explanatory variable of interest is the poverty rate, which is measured as the proportion of households in a district with an annual income below the federal poverty line. My outcome variable of interest is a U.S. Representative's DW-NOMINATE score (dimension 1), which measures a representative's economic ideology on a scale from -1 to 1, with -1 being the progressive end and 1 being the conservative end of the spectrum. These scores are calculated based on the representative's voting record in Congress. If I observe a negative correlation between a congressional district's poverty rate and its representative's DW-NOMINATE score, this would provide support for my hypothesis, since it would demonstrate that members of Congress representing districts with high poverty rates tend to be more economically progressive (i.e., have lower DW-NOMINATE scores). If, on the other hand, I observe a positive correlation between a district's poverty rate and its representative's DW-NOMINATE score, this would provide evidence against my hypothesis. When I run my regression of U.S. Representative DW-NOMINATE score on the congressional district poverty rate variable, a negative, significant coefficient would indicate support for my hypothesis. I will refer to an increase in the DW-NOMINATE score as an increase in U.S. Representative "economic conservatism."


## Data Analysis

```{r}

library(tidyverse)
congress_114 <- read_csv("H114_members.csv")
poverty_by_district <- read_csv("tab1_poverty_cd_2017.csv")

congress_114
poverty_by_district

district_poverty <- poverty_by_district |>
  select(district_id, poverty_rate)

district_poverty

```


```{r}

congress_and_poverty <- congress_114 |>
  left_join(district_poverty) 

congress_and_poverty

correlation <- cor(congress_and_poverty$poverty_rate, congress_and_poverty$nominate_dim1, use = "complete.obs")

knitr::kable(correlation,
             col.names = "Correlation Between  Poverty Rate and DW-NOMINATE Score", 
             digits = 2)
  
```

The correlation between poverty rate and DW-NOMINATE score is approximately -0.21. Members of Congress representing high-poverty districts are more likely to support ideologically progressive economic policies. This finding is consistent with my hypothesis.

```{r}

scatterplot_overall <- ggplot(congress_and_poverty, aes(x = poverty_rate, y = nominate_dim1)) +
  geom_point(color = "steelblue1") +
  labs(x = "Congressional District Poverty Rate",
       y = "DW-NOMINATE score (lower = more progressive)",
       title = "District Poverty Rate and U.S. Representative Economic Ideology") +
  geom_smooth(method = "lm", se = FALSE, color = "indianred1")

scatterplot_overall

```

## Linear Regression

```{r}

regression <- lm(nominate_dim1 ~ poverty_rate, data = congress_and_poverty)

regression

summary(regression)

regression2 <- lm(nominate_dim1 ~ poverty_rate + born, data = congress_and_poverty)

regression2

summary(regression2)

```

A 1-unit (100 percentage point) increase in the poverty rate of a congressional district is associated with a decrease in the U.S. Representative's DW-NOMINATE score of 1.877. A 1% increase in the poverty rate of a congressional district is associated with a decrease in the U.S. Representative's DW-NOMINATE score of 0.019. The relationship between congressional district poverty rate and U.S. Representative DW-NOMINATE score is statistically significant. 

## Analysis by Party

```{r}

dem_congress_and_poverty <- congress_and_poverty |>
  filter(party_code == 100)

rep_congress_and_poverty <- congress_and_poverty |>
  filter(party_code == 200)

dem_correlation <- cor(dem_congress_and_poverty$poverty_rate, dem_congress_and_poverty$nominate_dim1, use = "complete.obs")

knitr::kable(dem_correlation,
             digits = 2,
             col.names = "Correlation for Democrats")

scatterplot_dem <- ggplot(dem_congress_and_poverty, aes(x = poverty_rate, y = nominate_dim1)) +
  geom_point(color = "steelblue1") +
  labs(x = "Congressional District Poverty Rate",
       y = "DW-NOMINATE score (lower = more progressive)",
       title = "District Poverty Rate and Democrat Economic Ideology") +
  geom_smooth(method = "lm", se = FALSE, color = "indianred1")

scatterplot_dem

regression_dem <- lm(nominate_dim1 ~ poverty_rate, data = dem_congress_and_poverty)

regression_dem

summary(regression_dem)

rep_correlation <- cor(rep_congress_and_poverty$poverty_rate, rep_congress_and_poverty$nominate_dim1, use = "complete.obs")

scatterplot_rep <- ggplot(rep_congress_and_poverty, aes(x = poverty_rate, y = nominate_dim1)) +
  geom_point(color = "steelblue1") +
  labs(x = "Congressional District Poverty Rate",
       y = "DW-NOMINATE score (lower = more progressive)",
       title = "District Poverty Rate and Republican Economic Ideology") +
  geom_smooth(method = "lm", se = FALSE, color = "indianred1")

scatterplot_rep

knitr::kable(rep_correlation,
             digits = 2,
             col.names = "Correlation for Republicans")

regression_rep <- lm(nominate_dim1 ~ poverty_rate, data = rep_congress_and_poverty)

regression_rep

summary(regression_rep)

regression3 <- lm(nominate_dim1 ~ poverty_rate + party_code, data = congress_and_poverty)

regression3

summary(regression3)

```

We see that the correlation between congressional district poverty rate and U.S. Representative economic conservatism differs by party. Among Democrats, the correlation between congressional district poverty rate and U.S. Representative economic conservatism is -0.24, slightly stronger than the correlation for Congress overall. Consistent with my hypothesis, Democrats representing high-poverty districts are more likely to support progressive economic policies. However, among Republicans, the correlation between congressional district poverty rate and U.S. Representative economic conservatism is 0.05. Republicans representing high-poverty districts are slightly less likely to support progressive economic policies. This finding is inconsistent with my hypothesis, which predicts negative correlations between congressional district poverty rate and U.S. Representative economic conservatism. However, it is important to highlight that this correlation is weak.


## Final Report

# Introduction

To what extent is a U.S. congressional district's poverty rate correlated with its representative's economic ideology? Representatives with progressive economic ideologies are more likely to support policies -- such as anti-poverty programs or the redistribution of income to the poor -- that disproportionately benefit Americans living below the poverty line. However, a representative's economic ideology is also strongly influenced by political party affiliation. Democratic members of Congress are generally more supportive of progressive economic policies than their Republican counterparts. I hypothesize is that there is a negative correlation between a congressional district's poverty rate and its U.S. Representative's DW-NOMINATE score -- that is, congressional districts with higher rates of poverty are more likely to be represented by economic progressives. Furthermore, I hypothesize that this correlation holds true at the party level. In other words, I expect to see a negative correlation between congressional district poverty rate and economic conservatism among Republicans as well as among Democrats. What makes this hypothesis interesting is that many high-poverty districts in the United States, especially those located in rural areas, are represented by Republicans who tend to be averse to progressive economic policies that would disproportionately benefit their poor constituents. This report investigates how strong the correlation between congressional district poverty rate and representative economic ideology is in the U.S. Congress. The report will also run several regressions to assess the statistical significance of the relationship between congressional district poverty rate and representative economic ideology. 


# Data

The data referenced in this report was extracted from two sources: 1) the Food Research and Action Center's U.S. congressional district-level poverty data (based on the 2017 American Community Survey) and 2) the DW-NOMINATE (DW-N) scores for members of the 114th Congress (2015-17).

In this report, the independent variable is a U.S. congressional district's 2017 poverty rate, or the percentage of households in a district below the federal poverty line in 2017. According to the U.S. Department of Health and Human Services, a household of one was below the federal poverty line if that individual's income was less than $12,060 in 2017. A family of four was below the poverty line in 2017 if their household income was less than $24,600 that year. 

The dependent variable in this report is the DW-N score for a member of the 114th Congress. A DW-N score is calculated based on a U.S. Representative's voting record on economic issues while serving in Congress. A DW-N score closer to -1 is associated with economic progressivism (left-of-center), whereas a DW-N score closer to 1 is associated with economic conservatism (right-of-center). 

Below is a histogram that summarizing the dependent variable, DW-N score. The histogram illustrates the range of DW-N scores for members of the 114th Congress. Because the U.S. Congress is polarized along partisan lines, Democrats are clustered to the left of 0, whereas Republicans are clustered to the right of 0. 

```{r}

  histogram <- ggplot(data = congress_and_poverty, mapping = aes(x = nominate_dim1)) +
  geom_histogram() +
  labs(x = "DW-NOMINATE Score", y = "Number of U.S. Representatives",
       title = "Range of DW-NOMINATE Scores in 114th Congress")

histogram

```

The research design of this study is cross-sectional because it is analyzing the independent and dependent variables at one point in time. The poverty rates were calculated in 2017, and DW-N scores are held fixed for each member of the 114th Congress.


# Results

CORRELATION

What is the correlation between congressional district poverty rate and U.S. Representative economic ideology?

```{r}

correlation <- cor(congress_and_poverty$poverty_rate, congress_and_poverty$nominate_dim1, use = "complete.obs")

knitr::kable(correlation,
             col.names = "Correlation Between  Poverty Rate and DW-NOMINATE Score", 
             digits = 2)

scatterplot_overall <- ggplot(congress_and_poverty, aes(x = poverty_rate, y = nominate_dim1)) +
  geom_point(color = "steelblue1") +
  labs(x = "Congressional District Poverty Rate",
       y = "DW-NOMINATE score (lower = more progressive)",
       title = "District Poverty Rate and U.S. Representative Economic Ideology") +
  geom_smooth(method = "lm", se = FALSE, color = "indianred1")

scatterplot_overall

```

The correlation between poverty rate and DW-NOMINATE score is approximately -0.21. Members of Congress representing high-poverty districts are more likely to support ideologically progressive economic policies. This finding is consistent with my hypothesis.


LINEAR REGRESSION

```{r}

regression <- lm(nominate_dim1 ~ poverty_rate, data = congress_and_poverty)

regression

summary(regression)

```

A 1-unit (i.e., 100 percentage point) increase in the poverty rate of a congressional district is associated with a decrease in the U.S. Representative’s DW-NOMINATE score of 1.877. In other words, a 1 percentage point increase in the poverty rate of a congressional district is associated with a decrease in the U.S. Representative’s DW-NOMINATE score of 0.019. 

With a p-value of less than 0.001, the estimated coefficient for the poverty rate variable is highly statistically significant. However, it is not reasonable to conclude, based on this finding alone, that there is a causal effect between congressional district poverty rate and U.S. Representative DW-N score. To determine whether there is a causal effect, one must use multiple regression to control for potentially confounding variables.

One potential confounding variable is U.S. Representative year of birth. One plausible hypothesis is that older members of Congress are more economically conservative (have higher DW-N scores) because they are more likely to have grown up in a pre-welfare state America, when the federal government generally did not provide economic relief to the poor. How does the poverty rate coefficient change when adding U.S. Representative year of birth to the regression? Below is the result of running a linear regression of congressional district poverty rate on representative DW-N score, with the `born` variable included. 

```{r}

regression2 <- lm(nominate_dim1 ~ poverty_rate + born, data = congress_and_poverty)

regression2

summary(regression2)

```

Holding U.S. Representative birth year constant, a 1-unit (i.e., 100 percentage point) increase in the poverty rate of a congressional district is associated with a decrease in the U.S. Representative’s DW-NOMINATE score of 1.683. In other words, a 1 percentage point increase in the poverty rate of a congressional district is associated with a decrease in the U.S. Representative’s DW-NOMINATE score of 0.017.

The estimated coefficient for the `born` variable is highly statistically significant as well. Holding congressional district poverty rate constant, a 1-year increase in a representative's `born` year, or a 1-year decrease in representative age, predicts an increase in the DW-N score by 0.01. In other words, holding congressional district poverty rate constant, younger U.S. Representatives are significantly more likely to be economically conservative. This finding is inconsistent with the hypothesis that older U.S. Representatives are more economically conservative. 

With a p-value of less than 0.001, the estimated coefficient for the poverty rate variable is still highly statistically significant when controlling for representative birth year. Notwithstanding, it is unreasonable to conclude that there is a causal relationship between congressional district poverty rate and U.S. Representative DW-N score. To determine whether there is a causal relationship, one must also control for a crucial potential confounding variable: party affiliation. 

Below is the result of running the regression controlling for party affiliation and representative birth year. With a p-value of 0.15, the poverty rate coefficient is no longer statistically significant. This finding demonstrates that a congressional district's poverty rate likely does not have a causal effect on a U.S. Representative's economic ideology. 

Unsurprisingly, there is a highly statistically significant relationship between party affiliation and economic ideology. These results suggest that representatives' party affiliation very strongly influences their economic ideology as measured by the DW-N score. While this demonstrates that there is likely not a causal relationship between congressional district poverty rate and representative economic ideology, it does not change the fact that, as hypothesized, there is a negative correlation between congressional district poverty rate and DW-N score. 


```{r}

regression3 <- lm(nominate_dim1 ~ poverty_rate + born + party_code, data = congress_and_poverty)

regression3

summary(regression3)

```


ANALYSIS BY PARTY

How does the relationship between congressional district poverty rate and U.S. Representative economic ideology change when conducting the analysis party level? 

Below are scatterplots illustrating the correlations between congressional district poverty rate and U.S. Representative DW-N score for Democrats and Republicans. Below are also regressions run to analyze the significance of the relationship between congressional district poverty rate and representative economic ideology at the party level. 

```{r}

dem_correlation <- cor(dem_congress_and_poverty$poverty_rate, dem_congress_and_poverty$nominate_dim1, use = "complete.obs")

knitr::kable(dem_correlation,
             digits = 2,
             col.names = "Correlation for Democrats")

rep_correlation <- cor(rep_congress_and_poverty$poverty_rate, rep_congress_and_poverty$nominate_dim1, use = "complete.obs")

knitr::kable(rep_correlation,
             digits = 2,
             col.names = "Correlation for Republicans")

scatterplot_dem <- ggplot(dem_congress_and_poverty, aes(x = poverty_rate, y = nominate_dim1)) +
  geom_point(color = "steelblue1") +
  labs(x = "Congressional District Poverty Rate",
       y = "DW-NOMINATE score (lower = more progressive)",
       title = "District Poverty Rate and Democrat Economic Ideology") +
  geom_smooth(method = "lm", se = FALSE, color = "indianred1")

scatterplot_dem

scatterplot_rep <- ggplot(rep_congress_and_poverty, aes(x = poverty_rate, y = nominate_dim1)) +
  geom_point(color = "steelblue1") +
  labs(x = "Congressional District Poverty Rate",
       y = "DW-NOMINATE score (lower = more progressive)",
       title = "District Poverty Rate and Republican Economic Ideology") +
  geom_smooth(method = "lm", se = FALSE, color = "indianred1")

scatterplot_rep

regression_dem <- lm(nominate_dim1 ~ poverty_rate, data = dem_congress_and_poverty)

regression_dem

summary(regression_dem)

regression_rep <- lm(nominate_dim1 ~ poverty_rate, data = rep_congress_and_poverty)

regression_rep

summary(regression_rep)

```

The correlation between poverty rate and DW-N score differs markedly by party. Among Democrats, the correlation is -0.24, slightly stronger than the correlation for the 114th Congress as a whole. Running a regression between poverty rate and DW-N score among Democrats, the poverty rate coefficient is negative and statistically significant. Consistent with my hypothesis, Democrats representing high-poverty districts are more likely to support progressive economic policies.

Among Republicans, however, the correlation between congressional district poverty rate and representative DW-N score is in the opposite direction. The correlation is positive at 0.05, meaning that Republicans representing high-poverty districts are slightly less likely to support progressive economic policies. However, with a p-value of 0.454 for the poverty rate coefficient, this relationship is not statistically significant. Nonetheless, this finding is inconsistent with my hypothesis, which predicts a negative correlation between congressional district poverty rate and representative DW-N score for both parties. 


# Conclusion

I hypothesized that there is a negative correlation between congressional district poverty rate and U.S. Representative DW-N score -- that is, representatives of high-poverty districts are more likely to be economically progressive. Consistent with my hypothesis, the correlation was found to be -0.21. Moreover, the regression between these two variables is highly statistically significant. The relationship between these variables remained statistically significant when controlling for U.S. Representative year of birth. Surprisingly, there is a statistically significant negative relationship between representative age and DW-N score, meaning that older members of Congress are more likely to be economically progressive (holding congressional district poverty rate constant). As expected, when the political party affiliation variable is added to the regression, the relationship between congressional district poverty rate and representative DW-N score is no longer statistically significant. The likely reason is that representatives' economic ideologies are usually strongly influenced by their party affiliations. Therefore, one cannot infer a causal relationship between congressional district poverty rate and U.S. Representative economic ideology. Consistent with my hypothesis, there is a negative correlation between congressional district poverty rate and DW-N score among Democrats. However, contrary to my hypothesis, there is a slightly positive correlation between these two variables among Republicans. 

One limitation of my analysis is that it does not show that a congressional district's poverty rate has a causal effect on its representative's economic ideology. Although I did not hypothesize a causal relationship between these two variables, such a finding would have been more surprising and interesting because it would have suggested that members of Congress are willing to put partisanship aside to support the economic interests of their constituents. The negative correlations observed between these two variables may simply be due to the fact that there are more high-poverty Democratic districts than high-poverty Republican districts in the United States. My analysis could be improved in a future study by adding more variables to the regression to control for other potential confounding variables. This would require a larger dataset with more covariates than the one used in this analysis. In a future study, I believe it would strengthen the analysis to control for additional potential confounding variables, including race, gender identity, whether the representative grew up in poverty, and how long the representative has served in Congress. 

